---
title: "ImmunoTwin fullblood methylation analysis"
author: "Pablo Donner"
date: "2023-11-15"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(root.dir = "J:/DII ALL6/Projects/Pablo/November")
```

```{r}
# remove.packages("BiocManager")
# remove.packages("ExperimentHub")
# remove.packages("RPMM")
# remove.packages("pals")
# remove.packages("SummarizedExperiment")
# remove.packages("minfi")
# remove.packages("BiocParallel")
# remove.packages("GenomicRanges")
# remove.packages("sesame")
# remove.packages("sesameData")
# remove.packages("tibble")
# remove.packages("tidyr")
# remove.packages("gprofiler2")
# remove.packages("ggrepel")
# remove.packages("cowplot")
# remove.packages("corrplot")
# remove.packages("readxl")
```

#Packages
```{r}
##I recommend to uncomment this chunk as soon as all the packages are installed
# install.packages("pkgbuild")
# pkgbuild::check_build_tools()
# install.packages("devtools")
# require(devtools)
# install.packages("BiocManager")
# if (!require("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# BiocManager::install(version = "3.18")
# packageVersion("BiocManager")
# BiocManager::install("sesame")
# packageVersion("sesame")
# BiocManager::install("sesameData")
# packageVersion("sesameData")
# BiocManager::install("ExperimentHub")
# packageVersion("ExperimentHub")
# BiocManager::install("RPMM")
# BiocManager::install("pals")
# BiocManager::install("SummarizedExperiment")
# packageVersion("SummarizedExperiment")
# BiocManager::install("minfi")
# BiocManager::install("BiocParallel")
# packageVersion("BiocParallel")
# BiocManager::install("GenomicRanges")
# BiocManager::install("BiocFileCache")
# BiocManager::install("Bioconductor/BiocFileCache")
# install.packages("tibble")
# packageVersion("tibble")
# install.packages("tidyr")
# install.packages("gprofiler2")
# install.packages("ggrepel")
# install.packages("cowplot")
# install.packages("corrplot")
# install.packages("readxl")
# install.packages("stats")
# install.packages("reshape2")
install.packages("tidyverse")
```

```{r}
library(sesame)
library(sesameData)
library(ExperimentHub)
library(parallel)
library(ggplot2)
library(dplyr)
library(RPMM)
library(pals)
library(BiocFileCache)
packageVersion("BiocFileCache")
packageVersion("dbplyr")
library(SummarizedExperiment)
library(minfi)
library(BiocParallel)
library(GenomicRanges)
library(tibble)
library(tidyr)
library(gprofiler2)
library(ggrepel)
library(cowplot)
library(corrplot)
library(readxl)
library(stats)
library(reshape2)
library(tidyverse)
```


```{r}
#info 
packageVersion("sesame")
sessionInfo()
citation("sesame")
sesameDataCache()
```

##### MANIFEST (mft) #######

```{r}
# getting the manifest object
# mft = sesameDataGet("EPIC.address")$ordering
# head(mft, 20)

#Alternatively
tsv_path <- "J:/DII ALL6/Projects/Pablo/November/EPIC.hg38.manifest.tsv.gz"
sesameData_check_platform(platform = )
tsv_path
mft <- sesameAnno_buildAddressFile(tsv_path)
head(mft, 20)
```
#################### Opening IDAT files ###################

```{r}
#Import IDAT files, pair by pair and with all the IDATs in the same folder otherwise it doesn't work (error can't find GRN file)
#Do it in the right order (already clustered by group), simpler afterwards.
setwd("J:/DII ALL6/Projects/Pablo/November/Idat_Files")
sdf1 = readIDATpair("207717880162_R01C01", platform = "EPIC", manifest = mft, controls = NULL, verbose = TRUE)
sdf2 = readIDATpair("207717880162_R02C01", platform = "EPIC", manifest = mft, controls = NULL, verbose = TRUE)
sdf3 = readIDATpair("207717880162_R03C01", platform = "EPIC", manifest = mft, controls = NULL, verbose = TRUE)
sdf4 = readIDATpair("207717880162_R04C01", platform = "EPIC", manifest = mft, controls = NULL, verbose = TRUE)
sdf5 = readIDATpair("207717880162_R05C01", platform = "EPIC", manifest = mft, controls = NULL, verbose = TRUE)
sdf6 = readIDATpair("207717880162_R06C01", platform = "EPIC", manifest = mft, controls = NULL, verbose = TRUE)
sdf7 = readIDATpair("207717880162_R07C01", platform = "EPIC", manifest = mft, controls = NULL, verbose = TRUE)
sdf8 = readIDATpair("207717880162_R08C01", platform = "EPIC", manifest = mft, controls = NULL, verbose = TRUE)
sdf9 = readIDATpair("207717880174_R01C01", platform = "EPIC", manifest = mft, controls = NULL, verbose = TRUE)
sdf10 = readIDATpair("207717880174_R02C01", platform = "EPIC", manifest = mft, controls = NULL, verbose = TRUE)
sdf11 = readIDATpair("207717880174_R03C01", platform = "EPIC", manifest = mft, controls = NULL, verbose = TRUE)
sdf12 = readIDATpair("207717880174_R04C01", platform = "EPIC", manifest = mft, controls = NULL, verbose = TRUE)
sdf13 = readIDATpair("207717880174_R05C01", platform = "EPIC", manifest = mft, controls = NULL, verbose = TRUE)
sdf14 = readIDATpair("207717880174_R06C01", platform = "EPIC", manifest = mft, controls = NULL, verbose = TRUE)
sdf15 = readIDATpair("207717880174_R07C01", platform = "EPIC", manifest = mft, controls = NULL, verbose = TRUE)
sdf16 = readIDATpair("207717880174_R08C01", platform = "EPIC", manifest = mft, controls = NULL, verbose = TRUE)

#SigDF (sdf) objects are obtained
head(sdf1)    #example for sample1
```

################   PREPROCESSING   ##########--------------------------------------------------------------------------------------------

```{r}
#create a list of sdf files, regroup all the SigDFs together in a single list object
IDATs_november <- list(sdf1, sdf2, sdf3, sdf4, sdf5, sdf6, sdf7, sdf8,sdf9, sdf10, sdf11, sdf12, sdf13, sdf14, sdf15, sdf16)

# IDATs_november <- list(sdf1, sdf2)
# Assuming mft contains the manifest and IDATs_november is a list of sdf objects

# IDATs_november <- list(sdf1, sdf2)
#apply the openSesame() function to the list with the proper precprocessing code (here : QCDPB)
normalized_sdf_list_november <- openSesame(IDATs_november, prep = "QCDPB", prep_args = NULL, manifest = mft, func = NULL, BPPARAM = SerialParam(), platform = "EPIC")
normalized_sdf_list_november
```


```{r}
#check number of masked probes
lapply(IDATs_november, function(x) sum(x$mask))                            #number of masked probes before preprocessing
lapply(normalized_sdf_list_november, function(x) sum(x$mask))               #number of masked probes after preprocessing
```



```{r}
#visualisation of beta distribution
#for sample n°1 ([[1]])

par(mfrow=c(2,1), mar=c(3,3,2,1))
sesameQC_plotBetaByDesign(IDATs_november[[1]], main="Before", xlab="\beta")
rm(IDATs_november)#remove unecessary Data to seep up processing
sesameQC_plotBetaByDesign(matchDesign(normalized_sdf_list_november[[1]]), main="After", xlab="\beta")
```

########## Extracting Betas and inferring sample and mapping metadata  ################

#EXTRACTING BETAS and getting beta matrix
```{r}
#calculate beta values for every sample 
betas_list_november <- lapply(normalized_sdf_list_november, function(x) {getBetas(x, mask = TRUE)})
rm(normalized_sdf_list_november)

#combine betas from each sample into a single matrix
betas_november <- do.call("cbind", betas_list_november)
rm(betas_list_november)
head(betas_november)                #BETA MATRIX
dim(betas_november)
```

#Getting samples metadata
```{r}
#Creating a coldata df to infer sample metadata
setwd("C:/Program Files/RStudio/Code_and_data")
cd_november <- read.csv("C:/Program Files/RStudio/Code_and_data/coldatadf_november_clustered_Twins.csv")
# samplesheet
colnames(cd_november)
#[1] "IDAT"         "Sample_Group"   "Twins"
```

#Imputing sample names in the columns of the beta matrix
```{r}
cd_november$IDAT
colnames(betas_november) <- cd_november$IDAT
head(betas_november)
```

#Creating GRanges object from manifest (Getting probe metadata)
```{r}
#Download the manifest file from http://zwdzwd.github.io/InfiniumAnnotation#mouse and specify path
tsv_path <- "J:/DII ALL6/Projects/Pablo/November/EPIC.hg38.manifest.tsv.gz"

#create the genomic ranges (GRanges)  manifest object
manifest <- sesameAnno_buildManifestGRanges(tsv_path, genome="hg38", decoy=FALSE, columns=NULL)
#296070 probes in GRanges.
#7415 probes belong to chr*.
#0 probes on decoy chr.

#extract the correct probe order as character Vector from betas matrix
probe_ids_cv <- rownames(betas_november)          
head(probe_ids_cv)
#[1] "cg00101675_BC21" "cg00116289_BC21" "cg00211372_TC21" "cg00531009_BC21" "cg00747726_TC21"
#[6] "cg00896209_TC21"


#reorder manifest in right Probe_IDs order
manifest_reordered <- manifest[probe_ids_cv]
manifest_reordered
head(manifest_reordered)
view(manifest_reordered)
# GRanges object with 6 ranges and 0 metadata columns:
#              seqnames      ranges strand
#                 <Rle>   <IRanges>  <Rle>
#   cg26928153     chr1 10848-10849      +
#   cg16269199     chr1 10850-10851      +
#   cg13869341     chr1 15865-15866      -
#   cg14008030     chr1 18827-18828      -
#   cg12045430     chr1 29407-29408      -
#   cg20826792     chr1 29425-29426      -
#   -------
#   seqinfo: 26 sequences from an unspecified genome; no seqlengths


regs = sesameData_getTxnGRanges("hg38")
Probe_IDs = names(sesameData_getManifestGRanges("EPIC"))
anno = sesameData_annoProbes(Probe_IDs, promoters(regs), column="gene_name")


```

#Inferring metadata to betas matrix, creating a RangedSummarizedExperiment object
```{r}
#Creating a RangedSummarizedExperiment object 
Rse_november <- SummarizedExperiment(
  assay = betas_november,
  colData = cd_november,
  rowData = NULL,
  rowRanges = manifest_reordered,
  checkDimnames=TRUE)

Rse_november
```


#Checking coldata (Sample Metadata)
```{r}
#extract the relevant metadata from the column data slot of Rse object
# meta_november = dplyr::select(as_tibble(colData(Rse_november)), IDAT, Sample_Group, Twins)
# head(meta_november)
#check internal structure of meta object
# str(meta_november)              #the nominal variables are represented as characters 

# Re-assign the factors with levels 
colData(Rse_november)$Sample_Group <- factor(colData(Rse_november)$Sample_Group, levels = c("normal", "divergent"), ordered = FALSE)
colData(Rse_november)$Sample_Group <- relevel(colData(Rse_november)$Sample_Group, ref = "normal")

colData(Rse_november)$Twins <- factor(colData(Rse_november)$Twins, levels = c("1", "2", "3", "4", "5", "6", "7", "8"), ordered = FALSE)
#check for mistakes
str(colData(Rse_november))

meta_november = dplyr::select(as_tibble(colData(Rse_november)), IDAT, Sample_Group, Twins)

head(meta_november)
```


#filter out NAs (failed probes)
```{r}
#check if there is no NA
betas_november = assay(Rse_november)#extract beta matrix
head(betas_november)
ok1_november = checkLevels(betas_november, meta_november$Twins)
sum(ok1_november)     #gives probes that are ok    

betas_november[which(!ok1_november)[1],]       #gives the probes that failed


ok2_november = checkLevels(betas_november, meta_november$Sample_Group)
sum(ok2_november) #gives probes that are ok

#filter probes based on all indicators
filtered_betas_november = betas_november[ok1_november&ok2_november, ]
# rm(ok1_november)
# rm(ok2_november)
#shape of final betas
dim(betas_november)
dim(filtered_betas_november)
head(filtered_betas_november)
```


#checking for outliers in matched
```{r}

#create a boxplot for each matched Twinpair

plot_matched1 <- boxplot(
  abs(filtered_betas_november[, 1] - filtered_betas_november[, 2]),
    ylim = c(0, 1.1)  # Adjust the y-axis limits as needed
)

plot_matched1$stats

plot_matched2 <- boxplot(
  abs(filtered_betas_november[, 3] - filtered_betas_november[, 4]),
    ylim = c(0, 5)  # Adjust the y-axis limits as needed
)
plot_matched3 <- boxplot(
  abs(filtered_betas_november[, 5] - filtered_betas_november[, 6]),
    ylim = c(0, 5)  # Adjust the y-axis limits as needed
)
plot_matched4 <- boxplot(
  abs(filtered_betas_november[, 7] - filtered_betas_november[, 8]),
    ylim = c(0, 5)  # Adjust the y-axis limits as needed
)
plot_matched5 <- boxplot(
  abs(filtered_betas_november[, 9] - filtered_betas_november[, 10]),
    ylim = c(0, 5)  # Adjust the y-axis limits as needed
)
plot_matched6 <- boxplot(
  abs(filtered_betas_november[, 11] - filtered_betas_november[, 12]),
    ylim = c(0, 5)  # Adjust the y-axis limits as needed
)
plot_matched7 <- boxplot(
  abs(filtered_betas_november[, 13] - filtered_betas_november[, 14]),
    ylim = c(0, 5)  # Adjust the y-axis limits as needed
)
plot_matched8 <- boxplot(
  abs(filtered_betas_november[, 15] - filtered_betas_november[, 16]),
  ylim = c(0, 5)  # Adjust the y-axis limits as needed
)
# Extract outlier values
# Calculate the total sum of outliers for each plot_matched for the graph naming
sum_outliers_plot1 <- sum(length(plot_matched1$out))
sum_outliers_plot2 <- sum(length(plot_matched2$out))
sum_outliers_plot3 <- sum(length(plot_matched3$out))
sum_outliers_plot4 <- sum(length(plot_matched4$out))
sum_outliers_plot5 <- sum(length(plot_matched5$out))
sum_outliers_plot6 <- sum(length(plot_matched6$out))
sum_outliers_plot7 <- sum(length(plot_matched7$out))
sum_outliers_plot8 <- sum(length(plot_matched8$out))

# Print the total sum of outliers for each plot_matched
print(sum_outliers_plot1)
print(sum_outliers_plot2)
print(sum_outliers_plot3)
print(sum_outliers_plot4)
print(sum_outliers_plot5)
print(sum_outliers_plot6)
print(sum_outliers_plot7)
print(sum_outliers_plot8)

#total sum of outliers
total_outliers_matched <- sum(length(plot_matched1$out), length(plot_matched2$out), length(plot_matched3$out), length(plot_matched4$out),
                      length(plot_matched5$out), length(plot_matched6$out), length(plot_matched7$out), length(plot_matched8$out))

# Print the total sum of outliers for each row
print(total_outliers_matched)
```

#checking for outliers in random
```{r}

plot_random1 <- boxplot(
  abs(filtered_betas_november[, 11] - filtered_betas_november[, 6]),
    ylim = c(0, 5)  # Adjust the y-axis limits as needed
)
print(plot_random1)

plot_random2 <- boxplot(
  abs(filtered_betas_november[, 1] - filtered_betas_november[, 7]),
    ylim = c(0, 5)  # Adjust the y-axis limits as needed
)
plot_random3 <- boxplot(
  abs(filtered_betas_november[, 16] - filtered_betas_november[, 4]),
    ylim = c(0, 5)  # Adjust the y-axis limits as needed
)
plot_random4 <- boxplot(
  abs(filtered_betas_november[, 9] - filtered_betas_november[, 2]),
    ylim = c(0, 5)  # Adjust the y-axis limits as needed
)
plot_random5 <- boxplot(
  abs(filtered_betas_november[, 5] - filtered_betas_november[, 12]),
    ylim = c(0, 5)  # Adjust the y-axis limits as needed
)
plot_random6 <- boxplot(
  abs(filtered_betas_november[, 13] - filtered_betas_november[, 8]),
    ylim = c(0, 5)  # Adjust the y-axis limits as needed
)
plot_random7 <- boxplot(
  abs(filtered_betas_november[, 14] - filtered_betas_november[, 3]),
    ylim = c(0, 5)  # Adjust the y-axis limits as needed
)
plot_random8 <- boxplot(
  abs(filtered_betas_november[, 15] - filtered_betas_november[, 2]),
  ylim = c(0, 5)  # Adjust the y-axis limits as needed
)

# Extract outlier values
# Calculate the total sum of outliers for each plot_random
sum_outliers_plot1_random <- sum(length(plot_random1$out))
sum_outliers_plot2_random <- sum(length(plot_random2$out))
sum_outliers_plot3_random <- sum(length(plot_random3$out))
sum_outliers_plot4_random <- sum(length(plot_random4$out))
sum_outliers_plot5_random <- sum(length(plot_random5$out))
sum_outliers_plot6_random <- sum(length(plot_random6$out))
sum_outliers_plot7_random <- sum(length(plot_random7$out))
sum_outliers_plot8_random <- sum(length(plot_random8$out))

# Calculate the total sum of outliers for the random plots
total_outliers_random <- sum(
  length(plot_random1$out), length(plot_random2$out), length(plot_random3$out),
  length(plot_random4$out), length(plot_random5$out), length(plot_random6$out),
  length(plot_random7$out), length(plot_random8$out)
)

print(total_outliers_random)
#create total sum of outliers


```



```{r}

# Open a PDF file to save the plots
pdf("matched_and_random_plots_absolute_colored_with_names_and_counts_above_cutoff.pdf")

# Set colors for the boxplots

# Set names for the comparisons
plot_names_matched <- c(
  "1-2", "3-4", "5-6", "7-8", "9-10", "11-12", "13-14", "15-16"
)



# Code for matched plots with absolute values, colors for values above cutoff, names, and counts
plot_matched <- boxplot(
  abs(filtered_betas_november[, 1] - filtered_betas_november[, 2]),
  abs(filtered_betas_november[, 3] - filtered_betas_november[, 4]),
  abs(filtered_betas_november[, 5] - filtered_betas_november[, 6]),
  abs(filtered_betas_november[, 7] - filtered_betas_november[, 8]),
  abs(filtered_betas_november[, 9] - filtered_betas_november[, 10]),
  abs(filtered_betas_november[, 11] - filtered_betas_november[, 12]),
  abs(filtered_betas_november[, 13] - filtered_betas_november[, 14]),
  abs(filtered_betas_november[, 15] - filtered_betas_november[, 16]),
  names = plot_names_matched,  # Assign names
  ylim = c(0, 1.1)  # Adjust the y-axis limits as needed
)

# List of outliers sums for each column of x
outliers_sums <- c(sum_outliers_plot1, sum_outliers_plot2, sum_outliers_plot3, sum_outliers_plot4,
                   sum_outliers_plot5, sum_outliers_plot6, sum_outliers_plot7, sum_outliers_plot8)

text(
  x = 1:8,  # x-coordinates for text annotations (columns of x)
  y = apply(plot_matched$stats, 2, max) + 1,  # y-coordinate for text annotations (above each column)
  labels = outliers_sums,  # outlier sums per column of x
  pos = 3,  # position the text above the data points
  col = "red"  # color of the text
)
print(plot_matched)


# Code for random plots with absolute values, colors for values above cutoff, names, and counts
plot_names_random <- c(
  "6-11", "1-7", "16-4","9-2", "5-12", "13-8", "14-3", "15-2"
)
plot_random <- boxplot(
  abs(filtered_betas_november[, 11] - filtered_betas_november[, 6]),
  abs(filtered_betas_november[, 1] - filtered_betas_november[, 7]),
  abs(filtered_betas_november[, 16] - filtered_betas_november[, 4]),
  abs(filtered_betas_november[, 9] - filtered_betas_november[, 2]),
  abs(filtered_betas_november[, 5] - filtered_betas_november[, 12]),
  abs(filtered_betas_november[, 13] - filtered_betas_november[, 8]),
  abs(filtered_betas_november[, 14] - filtered_betas_november[, 3]),
  abs(filtered_betas_november[, 15] - filtered_betas_november[, 2]),
  names = plot_names_random,  # Assign names
  ylim = c(0, 1.1)  # Adjust the y-axis limits as needed
)

outliers_sums_random <- c(sum_outliers_plot1_random, sum_outliers_plot2_random, sum_outliers_plot3_random, sum_outliers_plot4_random,
                   sum_outliers_plot5_random, sum_outliers_plot6_random, sum_outliers_plot7_random, sum_outliers_plot8_random)
text(
  x = 1:8,  # x-coordinates for text annotations (columns of x)
  y = apply(plot_matched$stats, 2, max) + 1,  # y-coordinate for text annotations (above each column)
  labels = outliers_sums_random,  # outlier sums per column of x
  pos = 3,  # position the text above the data points
  col = "red"  # color of the text
)
print(plot_random)
# Save and close the PDF file
dev.off()

```


```{r}
# Calculate absolute differences
abs_diff <-abs(filtered_betas_november[, c(1, 3, 5, 7, 9, 11, 13, 15)] - filtered_betas_november[, c(2, 4, 6, 8, 10, 12, 14, 16)])


# Find rows where absolute differences are above 0.9
above_0.6_indices <- which(abs_diff > 0.6, arr.ind = TRUE)
rows_above_0.6 <- unique(above_0.6_indices[, 1])

# Extract CGs associated with these rows
selected_CpGs <- filtered_betas_november[rows_above_0.6, ]
head(selected_CpGs)
```

```{r}
# Find rows for each threshold level and store in respective matrices

above_0.1_indices <- which(abs_diff > 0.1, arr.ind = TRUE)
rows_above_0.1 <- unique(above_0.1_indices[, 1])
CpGs_Delta0.1 <- filtered_betas_november[rows_above_0.1, ]

above_0.2_indices <- which(abs_diff > 0.2, arr.ind = TRUE)
rows_above_0.2 <- unique(above_0.2_indices[, 1])
CpGs_Delta0.2 <- filtered_betas_november[rows_above_0.2, ]

above_0.3_indices <- which(abs_diff > 0.3, arr.ind = TRUE)
rows_above_0.3 <- unique(above_0.3_indices[, 1])
CpGs_Delta0.3 <- filtered_betas_november[rows_above_0.3, ]

above_0.4_indices <- which(abs_diff > 0.4, arr.ind = TRUE)
rows_above_0.4 <- unique(above_0.4_indices[, 1])
CpGs_Delta0.4 <- filtered_betas_november[rows_above_0.4, ]

above_0.5_indices <- which(abs_diff > 0.5, arr.ind = TRUE)
rows_above_0.5 <- unique(above_0.5_indices[, 1])
CpGs_Delta0.5 <- filtered_betas_november[rows_above_0.5, ]

above_0.6_indices <- which(abs_diff > 0.6, arr.ind = TRUE)
rows_above_0.6 <- unique(above_0.6_indices[, 1])
CpGs_Delta0.6 <- filtered_betas_november[rows_above_0.6, ]

above_0.7_indices <- which(abs_diff > 0.7, arr.ind = TRUE)
rows_above_0.7 <- unique(above_0.7_indices[, 1])
CpGs_Delta0.7 <- filtered_betas_november[rows_above_0.7, ]

above_0.8_indices <- which(abs_diff > 0.8, arr.ind = TRUE)
rows_above_0.8 <- unique(above_0.8_indices[, 1])
CpGs_Delta0.8 <- filtered_betas_november[rows_above_0.8, ]

above_0.9_indices <- which(abs_diff > 0.9, arr.ind = TRUE)
rows_above_0.9 <- unique(above_0.9_indices[, 1])
CpGs_Delta0.9 <- filtered_betas_november[rows_above_0.9, ]
```


```{r}

# Calculate absolute differences for all CpGs_Delta matrices
abs_diff_CpGs0.1 <- abs(CpGs_Delta0.1[, c(1, 3, 5, 7, 9, 11, 13, 15)] - CpGs_Delta0.1[, c(2, 4, 6, 8, 10, 12, 14, 16)])
abs_diff_CpGs0.2 <- abs(CpGs_Delta0.2[, c(1, 3, 5, 7, 9, 11, 13, 15)] - CpGs_Delta0.2[, c(2, 4, 6, 8, 10, 12, 14, 16)])
abs_diff_CpGs0.3 <- abs(CpGs_Delta0.3[, c(1, 3, 5, 7, 9, 11, 13, 15)] - CpGs_Delta0.3[, c(2, 4, 6, 8, 10, 12, 14, 16)])
abs_diff_CpGs0.4 <- abs(CpGs_Delta0.4[, c(1, 3, 5, 7, 9, 11, 13, 15)] - CpGs_Delta0.4[, c(2, 4, 6, 8, 10, 12, 14, 16)])
abs_diff_CpGs0.5 <- abs(CpGs_Delta0.5[, c(1, 3, 5, 7, 9, 11, 13, 15)] - CpGs_Delta0.5[, c(2, 4, 6, 8, 10, 12, 14, 16)])
abs_diff_CpGs0.6 <- abs(CpGs_Delta0.6[, c(1, 3, 5, 7, 9, 11, 13, 15)] - CpGs_Delta0.6[, c(2, 4, 6, 8, 10, 12, 14, 16)])
abs_diff_CpGs0.7 <- abs(CpGs_Delta0.7[, c(1, 3, 5, 7, 9, 11, 13, 15)] - CpGs_Delta0.7[, c(2, 4, 6, 8, 10, 12, 14, 16)])
abs_diff_CpGs0.8 <- abs(CpGs_Delta0.8[, c(1, 3, 5, 7, 9, 11, 13, 15)] - CpGs_Delta0.8[, c(2, 4, 6, 8, 10, 12, 14, 16)])
abs_diff_CpGs0.9 <- abs(CpGs_Delta0.9[, c(1, 3, 5, 7, 9, 11, 13, 15)] - CpGs_Delta0.9[, c(2, 4, 6, 8, 10, 12, 14, 16)])

# Displaying the head of abs_diff_CpGs0.9 as an example
head(abs_diff_CpGs0.9)
```

```{r}
# Find indices where values exceed the threshold
above_threshold_indices0.1 <- apply(abs_diff_CpGs0.1 > 0.1, 1, function(x) sum(x))
above_threshold_indices0.2 <- apply(abs_diff_CpGs0.2 > 0.2, 1, function(x) sum(x))
above_threshold_indices0.3 <- apply(abs_diff_CpGs0.3 > 0.3, 1, function(x) sum(x))
above_threshold_indices0.4 <- apply(abs_diff_CpGs0.4 > 0.4, 1, function(x) sum(x))
above_threshold_indices0.5 <- apply(abs_diff_CpGs0.5 > 0.5, 1, function(x) sum(x))
above_threshold_indices0.6 <- apply(abs_diff_CpGs0.6 > 0.6, 1, function(x) sum(x))
above_threshold_indices0.7 <- apply(abs_diff_CpGs0.7 > 0.7, 1, function(x) sum(x))
above_threshold_indices0.8 <- apply(abs_diff_CpGs0.8 > 0.8, 1, function(x) sum(x))
above_threshold_indices0.9 <- apply(abs_diff_CpGs0.9 > 0.9, 1, function(x) sum(x))

# Create a new column to store occurrences
abs_diff_hit_CpGs0.1 <- cbind(abs_diff_CpGs0.1, hits = above_threshold_indices0.1)
abs_diff_hit_CpGs0.2 <- cbind(abs_diff_CpGs0.2, hits = above_threshold_indices0.2)
abs_diff_hit_CpGs0.3 <- cbind(abs_diff_CpGs0.3, hits = above_threshold_indices0.3)
abs_diff_hit_CpGs0.4 <- cbind(abs_diff_CpGs0.4, hits = above_threshold_indices0.4)
abs_diff_hit_CpGs0.5 <- cbind(abs_diff_CpGs0.5, hits = above_threshold_indices0.5)
abs_diff_hit_CpGs0.6 <- cbind(abs_diff_CpGs0.6, hits = above_threshold_indices0.6)
abs_diff_hit_CpGs0.7 <- cbind(abs_diff_CpGs0.7, hits = above_threshold_indices0.7)
abs_diff_hit_CpGs0.8 <- cbind(abs_diff_CpGs0.8, hits = above_threshold_indices0.8)
abs_diff_hit_CpGs0.9 <- cbind(abs_diff_CpGs0.9, hits = above_threshold_indices0.9)
# Display the matrix with added column
view(abs_diff_hit_CpGs0.8)

```


```{r}

# out=tibble(threshold=NA,occour=NA,n=NA)

for (i in 1:9) {
  assign(paste0("abs_diff_hit_CpGs0.", i), get(paste0("abs_diff_hit_CpGs0.", i))[, "hits"])
}



table <- grep("abs_diff_hit_CpGs0", ls(), value = TRUE)

tmp <- get(table[1])
tmp_df <- data.frame(hits = tmp)

tmp <- tmp_df %>%
  group_by(hits) %>%
  summarise(n = n()) %>%
  mutate(threshold = hits) %>%
  glimpse()

out <- tmp

for(j in c(2:9)){
  tmp <- get(table[j])
  tmp_df <- data.frame(hits = tmp)
  tmp <- tmp_df %>%
    group_by(hits) %>% 
    summarise(n = n()) %>% 
    mutate(threshold=j) %>% 
    glimpse()
  
  out <- out %>% 
    add_row(tmp)
} 



```

```{r}

#infering gene from probe
sesameData_getGenesByProbes(
  c("cg02101808","cg20483857","cg05236660"),
  platform = "EPIC",
  genome = "hg38",
  max_distance = 100
)

#infering probe from gene  
sesameData_getProbesByGene(
gene_name = "SAMD15",
platform = "EPIC",
upstream = 0,
downstream = 0,
genome = "hg38"
)

ST6GAL21 <- sesameData_getProbesByGene(
gene_name = "ST6GAL2",
platform = "EPIC",
upstream = 0,
downstream = 0,
genome = "hg38"
)

names(ST6GAL21)
ST6GAL2_T <- as_tibble(ST6GAL21) %>% 
  mutate(CpGs=names(ST6GAL21)) %>% 
  glimpse()
view(ST6GAL21)

```

```{r}
genes_list <- read.csv("genes_list.csv")               #for protein coding genes only
gene_names <- genes_list$gene_name

#for all extracted genes 
lapply(gene_names, function(gene_name) {visualizeGene(gene_name, filtered_betas_november, 'EPIC')})
```


```{r}
visualizeGene("Tcf23", filtered_betas_november, 'EPIC')
```

###########   Differential Methylation Modeling   ###############-----------------------------------------------------------------------------------


###########################  EWAS   ##################################################### 

#linear modeling of the data (DML)

```{r}
#Differentially Methylatied Locus (DML)
# smry_november = DML(filtered_betas_november, ~Sample_Group+Sex, meta=meta_november, BPPARAM = SerialParam())            #(for two contrasts)


# sesameDataCache() # in case not done yet
# data <- sesameDataGet('HM450.76.TCGA.matched')
# smry <- DML(data$betas[1:1000,], ~type, meta=data$sampleInfo)
# 
# tmp<-data$sampleInfo
# 
# sesameDataGet_resetEnv()
# 
# 
# tmp<-meta_november

row.names(meta_november) <- meta_november$IDAT
head(meta_november)
filtered_betas_november_top <- filtered_betas_november[1:10000,]

smry_november = DML(filtered_betas_november_top, ~Sample_Group, meta=meta_november, BPPARAM = SerialParam())
smry_november

smry_november[[1]]                                  #for first sample

res_november = summaryExtractTest(smry_november)#extract results
rm(smry_november)
head(res_november)
dim(res_november) 
str(res_november)
#size of dataframe
#[1] 272337     7
```
###Rename for pivoting###
```{r}
names(res_november)[names(res_november) == "Pval_Sample_Groupnormal"] <- "Pval_Twins1"
names(res_november)[names(res_november) == "Est_Sample_Groupnormal"] <- "Est_Twins1"
head(res_november)
```
```{r}
subset_res_november <- res_november[res_november$Pval_Sample_Groupnormal < 0.05, ]
subset_res_november

subset_res_november_1 <- res_november[res_november$Pval_Sample_Groupnormal < 0.01, ]
subset_res_november_1

sig_probe_ID_list_1 <- subset_res_november$Probe_ID                   ##used the <0.01 threshold for now (since unadjusted)
# Remove probes not in res_november
sig_probe_ID_list_1 <- sig_probe_ID_list_1[sig_probe_ID_list_1 %in% res_november$Probe_ID]

# Subset the beta matrix to include only the rows with probe IDs in the list
subset_beta_matrix <- filtered_betas_november[rownames(filtered_betas_november) %in% sig_probe_ID_list_1, ]

head(subset_beta_matrix)
```

```{r}
setwd("J:/DII ALL6/Projects/Pablo/November/Idat_Files")
sdf1 = readIDATpair("207717880162_R01C01", platform = "EPIC", manifest = mft, controls = NULL, verbose = TRUE)
sdf2 = readIDATpair("207717880162_R02C01", platform = "EPIC", manifest = mft, controls = NULL, verbose = TRUE)
sdf3 = readIDATpair("207717880162_R03C01", platform = "EPIC", manifest = mft, controls = NULL, verbose = TRUE)
sdf4 = readIDATpair("207717880162_R04C01", platform = "EPIC", manifest = mft, controls = NULL, verbose = TRUE)
sdf5 = readIDATpair("207717880162_R05C01", platform = "EPIC", manifest = mft, controls = NULL, verbose = TRUE)
sdf6 = readIDATpair("207717880162_R06C01", platform = "EPIC", manifest = mft, controls = NULL, verbose = TRUE)
sdf7 = readIDATpair("207717880162_R07C01", platform = "EPIC", manifest = mft, controls = NULL, verbose = TRUE)
sdf8 = readIDATpair("207717880162_R08C01", platform = "EPIC", manifest = mft, controls = NULL, verbose = TRUE)
sdf9 = readIDATpair("207717880174_R01C01", platform = "EPIC", manifest = mft, controls = NULL, verbose = TRUE)
sdf10 = readIDATpair("207717880174_R02C01", platform = "EPIC", manifest = mft, controls = NULL, verbose = TRUE)
sdf11 = readIDATpair("207717880174_R03C01", platform = "EPIC", manifest = mft, controls = NULL, verbose = TRUE)
sdf12 = readIDATpair("207717880174_R04C01", platform = "EPIC", manifest = mft, controls = NULL, verbose = TRUE)
sdf13 = readIDATpair("207717880174_R05C01", platform = "EPIC", manifest = mft, controls = NULL, verbose = TRUE)
sdf14 = readIDATpair("207717880174_R06C01", platform = "EPIC", manifest = mft, controls = NULL, verbose = TRUE)
sdf15 = readIDATpair("207717880174_R07C01", platform = "EPIC", manifest = mft, controls = NULL, verbose = TRUE)
sdf16 = readIDATpair("207717880174_R08C01", platform = "EPIC", manifest = mft, controls = NULL, verbose = TRUE)

# Assuming 'res_november_no_missing' is your data frame and 'IDAT' is a column containing IDAT values
Twins <- c("207717880162_R01C01", "207717880162_R02C01", "207717880162_R03C01", "207717880162_R04C01", "207717880162_R05C01", "207717880162_R06C01", "207717880162_R07C01", "207717880162_R08C01", "207717880174_R01C01", "207717880174_R02C01", "207717880174_R03C01", "207717880174_R04C01", "207717880174_R05C01", "207717880174_R06C01", "207717880174_R07C01", "207717880174_R08C01")  # List of specific IDAT values

# Create pairs of IDAT values
Twin_pairs <- combn(Twins, 2, simplify = TRUE)

# Loop through each pair of IDAT values and create paired plots
for (i in 1:ncol(Twin_pairs)) {
  # Filter data for the specific pair of IDAT values
  filtered_data <- res_november_no_missing %>%
    filter(IDAT %in% Twin_pairs[, i])
  
  # Create ggplot for the filtered data
  plot_title <- paste("Plot for Twin pairs:", paste(Twin_pairs[, i], collapse = " and "))  # Create a title for each pair
  plot <- ggplot(filtered_data) +
    geom_point(aes(x = Est_Sample_Groupnormal, y = -log10(Pval_Sample_Groupnormal))) +
    coord_cartesian(ylim = c(0, 7)) +
    ggtitle(plot_title)  # Set the plot title
  
  # Print each plot
  print(plot)
}

res_november_no_missing <- na.omit(res_november)    
#ask whether DNA methylation is different between H1N1 and Ctl groups
#Est_Sample_GroupH1N1 compares H1N1 group to Ctl
ggplot(res_november_no_missing) + geom_point(aes(Est_Sample_Groupnormal, -log10(Pval_Sample_Groupnormal))) +
  coord_cartesian(ylim = c(0, 7))
```

###Reshape df to long format based on cg###
```{r}
# df_long <- df %>%
#   pivot_longer(cols = cg, names_to = Probe_ID, values_to = "Probe_ID")
# head(df_long)

names(df)[names(df) == "cg"] <- "Probe_ID"

res_november_org <- res_november %>% 
  pivot_longer(c(Est_Twins1,Est_Twins2,Est_Twins3,Est_Twins4,Est_Twins5,Est_Twins6, Est_Twins7, Est_Twins8), values_to = "EST",names_to = "Twin") %>% 
  pivot_longer(c(Pval_Twins1,Pval_Twins2,Pval_Twins3,Pval_Twins4,Pval_Twins5,Pval_Twins6,Pval_Twins7, Pval_Twins8), values_to = "Pval",names_to = "Twin2") %>% 


# Join the reshaped datasets

df_sig <- res_november_org %>% 
  pivot_longer(contains("Twins"),values_to = "value",names_to = "Twin") %>% 
  separate_wider_delim(Twin,delim = "_",names= c("type", "Twin")) %>%
  pivot_wider(names_from = type,values_from = value) %>% 
  separate_wider_delim(Twin,delim = "Twins",names = c(NA,"Twins")) %>% 
  # pivot_longer(contains("Pval_Twins"),values_to = "Pval",names_to = "Twin2") %>% 
 
  # pivot_longer(contains("Pval_Twins"),values_to = "Pval",names_to = "Twin2") %>% 
  

# Join the reshaped datasets by Probe_ID
df_sig2 <- df %>%
  full_join(df_sig) %>% 
  head()

```




###CHECK IF A TWINPAIR HAS NO SIGNIFICANT METHYLATION###
```{r}


twin_pairs <- df_sig2 %>%
  filter(rowSums(select(., starts_with("Pval"))) < 0.05) %>%
  group_by(Twins) %>%
  filter(n() > 1) %>%
  ungroup()

for (pair in 1:8) {
  df_filtered <- twin_pairs %>%
    filter(Probe_ID %in% cg_values) %>%
    filter(Pval < 0.05) %>%
    filter(Twins == pair)  # Assuming Twins corresponds to the version

  if (nrow(df_filtered) > 0) {
    print(paste("There are values meeting the condition for Twin Pair", pair))
  } else {
    print(paste("No values meeting the condition for Twin Pair", pair))
  }
}

```

#trial
```{r}

 # Assigning diffexpressed 
res_november$diffexpressed <- "NO"
res_november$diffexpressed[res_november$Est_Twins > 0.6 & -log10(res_november$Pval_Twins) < 0.05] <- "UP"
res_november$diffexpressed[res_november$Est_Twins < -0.6 & -log10(res_november$Pval_Twins) < 0.05] <- "DOWN" 


# Filtering the dataframe and selecting the first n rows
res_november_filtered <- res_november %>%
  filter(n() > 1) %>%
   filter(!is.na(Pval_Twins1)) %>%
  # filter(Twins %in% c("1", "5"))%>%
  filter(row_number() <= 10000)%>%
  ungroup()%>%
  glimpse()

mycolors <- c("green", "red", "black")
names(mycolors) <- c("DOWN", "UP", "NO")

# Generating volcano plots for Twin Pair 1 (considering only the first 100 rows)
volcano_plot <- ggplot(res_november_filtered, aes(x = Est, y = -log10(Pval), col = diffexpressed)) +
  geom_point() +
  geom_vline(xintercept = c(-0.6, 0.6), col = "red") +
  geom_hline(yintercept = -log10(0.05), col = "red") +
  coord_cartesian(ylim = c(0, 7)) +
  ggtitle("Volcano Plot") +
  labs(x = "Delta Beta", y = "-log10(P-value)") +
  # facet_wrap(~Twins) +
  theme_classic() +
  scale_color_manual(values = mycolors) 
  # geom_text(hjust = -0.1, vjust = -0.5, size = 3, show.legend = FALSE)

print(volcano_plot)




```


####Volcano Plot###
```{r}

twin_pairs <- df_sig %>%
  filter(rowSums(across(starts_with("Pval_Twins")) < 0.05) > 0) %>%
  group_by(Twins) %>%
  filter(n() > 1) %>%
  ungroup()

# Generating volcano plots for each twin pair
for (pair in unique(twin_pairs$Twins)) 
  if(pair == 1) {
  df_filtered <- twin_pairs %>%
    filter(Twins == pair) %>%
    na.omit()  # Removing rows with missing values

  volcano_plot <- ggplot(df_filtered, aes(x = get(paste0("Est_Twins", pair)), y = -log10(get(paste0("Pval_Twins", pair))))) +
    geom_point() +
    geom_vline(xintercept = c(-0.6, 0.6), col = "red") +  # Vertical lines at -0.6 and 0.6
    geom_hline(yintercept = -log10(0.05), col = "red") +  # Horizontal line at -log10(0.05)
    coord_cartesian(ylim = c(0, 7)) +
    ggtitle(paste("Volcano Plot for Twin Pair1", pair)) +
    labs(x = "log2fc", y = "-log10(P-value)") +
    theme_classic()
  
  print(volcano_plot)
}


df_sig$diffexpressed <- "NO"
df_sig$diffexpressed[df_sig[[paste0("Est_Twins", pair)]] > 0.6 & df_sig[[paste0("Pval_Twins", pair)]] < 0.05] <- "UP"
df_sig$diffexpressed[df_sig[[paste0("Est_Twins", pair)]] < -0.6 & df_sig[[paste0("Pval_Twins", pair)]] < 0.05] <- "DOWN"

# Re-plotting with color based on 'diffexpressed'
volcano_plot <- ggplot(data = df_sig, aes(x = get(paste0("Est_Twins", pair)), y = -log10(get(paste0("Pval_Twins", pair))), col = diffexpressed)) +
  geom_point() +
  theme_minimal()

# Adding lines as specified
volcano_plot <- volcano_plot + geom_vline(xintercept = c(-0.6, 0.6), col = "red") +
  geom_hline(yintercept = -log10(0.05), col = "red")

print(volcano_plot)


```  

###create PDF for Volcano Plots###
```{r}    
pdf("all_volcano_plots.pdf")

for (pair in unique(twin_pairs$Twins)) {
  df_filtered <- twin_pairs %>%
    filter(Twins == pair) %>%
    na.omit()  # Removing rows with missing values
  
  print(paste("Processing Twin Pair:", pair))
  print(head(df_filtered))  # Print the first few rows of the filtered data for debugging
  
  volcano_plot <- ggplot(df_filtered, aes(x = get(paste0("Est_Twins", pair)), y = -log10(get(paste0("Pval_Twins", pair))))) +
    geom_point(aes(color = Sample_Group)) +
    coord_cartesian(ylim = c(0, 7)) +
    ggtitle(paste("Volcano Plot for Twin Pair", pair)) +
    labs(x = "-log10(P-value)", y = "betas") +
    theme_classic()

  print(volcano_plot)
}

dev.off()

```

###Scatterplots for certain CPGs###
```{r}


  cg_values <- c('cg00000155', 'cg00000596', 'cg00000622', 'cg00000658', 'cg00000714') # Define the specific cg values of interest

# Creating separate plots for each version of Pval_Twins
scatter_plots <- lapply(1:8) function(version) {
  df_filtered <- df_sig %>%
    filter(Probe_ID %in% cg_values) %>%
    filter(get(paste0("Pval_Twins", version)) < 0.05) %>%
    filter(Twins == version)  # Assuming Twins corresponds to the version


# Combine plots into one plot object
multiplot_scatter <- do.call(gridExtra::grid.arrange, c(scatter_plots, ncol = 2))  # Adjust ncol as needed

# Show the combined plot
print(multiplot_scatter)
print(scatter_plots[[1]])
print(scatter_plots[[3]])
print(scatter_plots[[5]])
print(scatter_plots[[7]])
```

###create PDF for Scatterplots###
```{r}
# Create a PDF file for all plots (multiplot)
pdf("all_scatter_plots.pdf")
print(multiplot_scatter)
dev.off()

# Save individual plots to separate PDF files
pdf("scatter_plot_1.pdf")
print(scatter_plots[[1]])
dev.off()

pdf("scatter_plot_3.pdf")
print(scatter_plots[[3]])
dev.off()

pdf("scatter_plot_5.pdf")
print(scatter_plots[[5]])
dev.off()

pdf("scatter_plot_7.pdf")
print(scatter_plots[[7]])
}

dev.off()
```


################# Manhattan plot ############################
```{r}

#Extract probe ID and p-value
probe_ID <- res_november$Probe_ID
pval <- res_november$Pval_Sample_Groupdivergent

# Create a new data frame with probe ID and p-value
pval_vector <- setNames(as.numeric(pval), probe_ID)

head(pval_vector)

#visualize the chromosomal distribution and significance level of CpG probes from an EWAS study
#manhattan plot
KYCG_plotManhattan(-log10(pval_vector), platform="EPIC",
    col=c("darkgreen","red"), ylabel = bquote(-log[10](P~value)), label_min=30) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  coord_cartesian(ylim = c(0, 7))
```

```{r}
pdf("manhattan plot.pdf", width = 7, height = 4)
KYCG_plotManhattan(-log10(pval_vector), platform="EPIC",
col=c("darkgreen","red"), ylabel = bquote(-log[10](P~value)), label_min=30) +
theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
coord_cartesian(ylim = c(0, 7))

dev.off()
```


